---
title: "J15221 Evanthia TNBC - Correlation of Luminex with response and subtype"
author: "Ludmila Danilova & Joseph Tandurella"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(data.table)
library("DT")
library(gridExtra)
library(scales)

##Create All Facet for plotting
CreateAllFacet <- function(df, col){
  df$facet <- df[[col]]
  temp <- df
  temp$facet <- "All"
  merged <-rbind(temp, df)

  # ensure the facet value is a factor
  merged[[col]] <- as.factor(merged[[col]])

  return(merged)
}


```

## Create figures for the clinical paper
Luminex data units:
ng/ml	for sCD137 and	Granzyme A

pg/ml	for IFNg	IL-10	IL-12(p40)	IL-12(p70)	IL-2	IL-6	IL-8


```{r  fig.width = 10}
# labels for time points
timeLab  = c("Baseline", "Post Run-in", "Week 8")

# set colors
recistCol = c("Non-responder" = "red","Responder" = "blue")
cbrCol = c("CR,PR,SD" = "blue","PD" = "red")
subtypeCol = c("HR+" = "red", "TNBC" = "blue")
timeCol = setNames(c("black", 'grey45', 'grey75'), timeLab)

# set root folder
rootDir = "./"

# Load data patient data
dta_pts = read.csv(paste0(rootDir,"input_tables/table_patient_response_OS.csv"), row.names = 1)
dta_pts$ID = rownames(dta_pts)

#==========================
# file with TILs and PDL1 data
dat = read.csv(paste0(rootDir,"input_tables/table_per_sample_values.csv"))


#===========================
# add patient info 
dat = cbind(dat,  dta_pts[dat$patientID,])

df_Baseline <- dat %>% filter(timePoint == 'Baseline')

df_sampInfo <- CreateAllFacet(dat, "Subtype")
df_sampInfo_recist <- CreateAllFacet(dat, "RECIST")

cytokines = colnames(dat)[10:18]
units = setNames(rep(c("pg/ml", "ng/ml"), times = c(7,2)), cytokines)

```


## Baseline samples. Cytokines by subtype 


```{r }

#================


##Type (HR+ vs TNBC)
for(i in cytokines)
{
  df = cbind(protein = df_Baseline[,i], df_Baseline[,c("RECIST","Subtype") ] )
  print(ggplot(df, aes(x=Subtype, y=protein)) + 
    geom_boxplot() + 
    labs(title=paste0(i,". Baseline Samples")) + 
    stat_compare_means(comparisons= list(c("HR+","TNBC")), label.y.npc = 0.85, method = "wilcox")+ 
    scale_y_continuous(trans = "log10", labels = comma) +
    theme_bw()+ 
    ylab(paste(i,"(",units[i],")"))+
    geom_point(shape=16))
}
```


## Cytokines by RECIST 
### Table 
with mean and +/- standard deviation for responders and non-responder by RECIST, and p-value from Wilcoxon rank-sum test

Additionally, Wilcoxon test for each time point comparing responders vs non-responders. And paired Wilcoxon p-values for baseline vs post run-in and post run-in vs week 8.
Then take differences between all time points and compare responders vs non-responders to see if there is significant difference in changes after treatment


```{r echo=FALSE,  fig.width = 10}

# create table with descriptive statistics for responders and non-responders and p-values

tab = sapply(cytokines, function(i){
  df = data.frame(protein = dat[,i],dat[,c("timePoint","RECIST")]) 
  # responders
  resp = df %>% filter(RECIST == "Responder" & timePoint == timeLab[1]) %>% select(protein) %>% unlist()
  resp1 = df %>% filter(RECIST == "Responder" & timePoint == timeLab[2]) %>% select(protein) %>% unlist()
  resp2 = df %>% filter(RECIST == "Responder" & timePoint == timeLab[3]) %>% select(protein) %>% unlist()
  # non-responders
  nresp = df %>% filter(RECIST == "Non-responder" & timePoint == timeLab[1])%>% select(protein)%>% unlist()
  nresp1 = df %>% filter(RECIST == "Non-responder" & timePoint == timeLab[2])%>% select(protein)%>% unlist()
  nresp2 = df %>% filter(RECIST == "Non-responder" & timePoint == timeLab[3])%>% select(protein)%>% unlist()

  # p-values
   res = compare_means(protein ~ RECIST, df, group.by = "timePoint")
   
 c(baseline_mean_resp = round(mean(resp, na.rm = T),2),
   baseline_mean_non_resp = round(mean(nresp, na.rm = T),2),
   baseline_p_value = res[1,'p.format'], 
   
   timepoint1_mean_resp = round(mean(resp1, na.rm = T),2),
   timepoint1_mean_non_resp = round(mean(nresp1, na.rm = T),2),
   timepoint1_p_value = res[2,'p.format'],
   
   timepoint2_mean_resp = round(mean(resp2, na.rm = T),2),
   timepoint2_mean_non_resp = round(mean(nresp2, na.rm = T),2),
   timepoint2_p_value = res[3,'p.format']) 

})

tab = t(tab)
colnames(tab) = gsub(".p.format","", colnames(tab))


datatable(tab)


res = c()
  for (k in cytokines)
  {
    mat <- data.frame(matrix(nrow = length(unique(dat$patientID)),
      ncol = 3, dimnames = list(unique(dat$patientID),
                      levels(factor(dat$timePoint)))), check.names = F)
    for(i in levels(factor(dat$timePoint)))
    {
      d1 = dat %>% filter(timePoint == i)
      mat[d1$patientID, i] = d1[,k]
    }
    # add response
    mat$RECIST = dta_pts[rownames(mat),"RECIST"]
    # create treatment effect 
    # substract baseline from post run-in and week 8 
    mat$delta_T1_B =  mat[,2]-mat[,1]
    mat$delta_T2_B =  mat[,3]-mat[,1]
    # and week 8 from post run-in 
    mat$delta_T2_T1 =  mat[,3]-mat[,2]
    
    # running paired test
    # Baseline vs post run-in
    p1 = wilcox.test(mat[,1], mat[,2], paired = T)$p.value
    # post run-in vs week 8
    p2 = wilcox.test(mat[,2], mat[,3], paired = T)$p.value
    
    # test by response
    # for baseline 
    p3 = wilcox.test(Baseline ~ RECIST, data = mat)$p.value
    # for post run-in 
    p4 = wilcox.test(mat[,2] ~ mat[,"RECIST"], data = mat)$p.value
    # for week 8
    p5 = wilcox.test(mat[,3] ~ mat[,"RECIST"], data = mat)$p.value
    # compare differences in treatment effect in responders vs non-responders
    p6 = wilcox.test(delta_T1_B ~ RECIST, data = mat)$p.value
    p7 = wilcox.test(delta_T2_B ~ RECIST, data = mat)$p.value
    p8 = wilcox.test(delta_T2_T1 ~ RECIST, data = mat)$p.value
   
    res = rbind(res, c(BvsT1 = p1,T1vsT2 = p2, B_RvsNR = p3, T1_RvsNR = p4, T2_RvsNR = p5, d_T1_B_RvsNR = p6, d_T2_B_RvsNR = p7, d_T2_T1_RvsNR = p8))
  }

rownames(res) = cytokines
res = round(res,3)

datatable(res)


```

### Boxplots of all samples by RECIST and time point

```{r fig.width = 10}

##RECIST (Responders vs. Nonresponders)
for(i in cytokines)
{
  df = cbind(protein = dat[,i], dat[,c("RECIST","Subtype", "timePoint") ] )
print(ggplot(df, aes(x=RECIST, y=protein)) +
  geom_boxplot() +
  labs(title=paste0(i,". All Samples")) +
  ylab(paste(i,"(",units[i],")"))+
  stat_compare_means(method = "wilcox.test") +
  geom_point(aes(color=Subtype)) +
  scale_y_log10()+
  scale_color_manual(values = subtypeCol) +
  facet_wrap("timePoint")+
  theme_bw())
}
## paired boxplots by time and RECIST
for(i in cytokines[1:9])
{
  df = cbind(protein = df_sampInfo_recist[,i], df_sampInfo_recist[,c("patientID", "timePoint", "RECIST","Subtype", "facet") ] )
  
    print(ggpaired(df , x="timePoint",y="protein", id = "patientID", line.size = 0.4, line.color = "RECIST",palette =  recistCol,
     title = paste0(i, ". All Samples"), xlab="Time", ylab = paste(i,"(",units[i],")")) +
      facet_wrap("facet") +
      scale_x_discrete(labels=timeLab)+
      scale_y_continuous(trans = "log10", labels = comma)+
      stat_compare_means(comparisons = list(c(timeLab[1],timeLab[2]), c(timeLab[1],timeLab[3]), c(timeLab[2],timeLab[3]))))
}

```

### Boxplots of baseline samples by RECIST and subtype

```{r   fig.width = 10}

##RECIST (Responders vs. Nonresponders)
for(i in cytokines)
{
  df = cbind(protein = df_Baseline[,i], df_Baseline[,c("RECIST","Subtype") ] )
print(ggplot(CreateAllFacet(df, "Subtype"), aes(x=RECIST, y=protein)) +
  geom_boxplot() +
  labs(title=paste0(i,". Baseline Samples")) +
  ylab(paste(i,"(",units[i],")"))+
  stat_compare_means() +
  geom_point(aes(color=Subtype)) +
  scale_y_log10()+
  scale_color_manual(values = subtypeCol) +
  facet_wrap("facet")+
  theme_bw())
}
# create table with descriptive statistics for responders and non-responders and p-values
tab = sapply(cytokines, function(i){
  df = data.frame(protein = df_Baseline[,i],RECIST = df_Baseline[,"RECIST"] )
  # responders
  resp = df %>% filter(RECIST == "Responder")
  # non-responders
  nresp = df %>% filter(RECIST == "Non-responder")
 
 c(unit = units[i], 
   mean_resp = round(mean(resp$protein, na.rm = T),2),
   sd_resp = round(sd(resp$protein, na.rm = T),2),
   mean_non_resp = round(mean(nresp$protein, na.rm = T),2),
   sd_non_resp = round(sd(nresp$protein, na.rm = T),2),
   p_value = compare_means(protein ~ RECIST, df)$p.format) #p-values
})


```


## Cytokines by subtype and RECIST


```{r  fig.width = 10}

##by type and RECIST
for(i in cytokines[1:9])
{
  df = cbind(protein = df_sampInfo[,i], df_sampInfo[,c("patientID", "timePoint", "RECIST","Subtype", "facet") ] )
  
    print(ggpaired(df , x="timePoint",y="protein", id = "patientID", line.color = "RECIST", line.size = 0.4,
     palette =  recistCol, title = paste0(i, ". All Samples"), xlab="Time", ylab = paste(i,"(",units[i],")")) +
      scale_x_discrete(labels=timeLab)+
      facet_wrap("facet") +
      scale_y_continuous(trans = "log10", labels = comma)+
      stat_compare_means(comparisons = list(c(timeLab[1],timeLab[2]), c(timeLab[1],timeLab[3]), c(timeLab[2],timeLab[3]))))
}


```
```{r }
sessionInfo()
```




