---
title: "J15221 Evanthia TNBC - CyTOF marker expression"
author: "Ludmila Danilova & Joseph Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(data.table)
library("DT")
library(gridExtra)
library(scales)

source("functions.r")
```

Create figures for the clinical paper

All data is coming from tables for the paper


```{r fig.width = 10}
# labels for time points
timeLab  = c("Baseline", "Post Run-in", "Week 8")

# set colors
recistCol = c("Non-responder" = "red","Responder" = "blue")
cbrCol = c("CR,PR,SD" = "blue","PD" = "red")
subtypeCol = c("HR+" = "red", "TNBC" = "blue")
timeCol = setNames(c("black", 'grey45', 'grey75'), timeLab)

# set root folder
rootDir = "./"

# Load data patient data
dta_pts = read.csv(paste0(rootDir,"input_tables/table_24patient_info.csv"), row.names = 1)
dta_pts$ID = rownames(dta_pts)

# remove patients with no response
dta_pts = dta_pts %>% filter(Best_Resp != "Unknown")
dim(dta_pts)
table(dta_pts$RECIST)

#==========================
# file with TILs and PDL1 data
cytof_dat = read.csv(paste0(rootDir,"input_tables/table_cytof_marker_expression.csv"))


markers = colnames(cytof_dat)[4:ncol(cytof_dat)]
#===========================
# add patient info 
cytof_dat = cbind(cytof_dat,  dta_pts[cytof_dat$patientID,])
cat("The number of patients with CYTOF data:", length(intersect(cytof_dat$patientID, dta_pts$ID)))
# print the number of pairs between time points
mat = printPairs(cytof_dat)

# take only baseline and time 1 values
df_BvT1 <- cytof_dat %>% filter(timePoint %in% c('Baseline','Post Run-in'))

# take only time 1 and time 2 values
df_T1vT2 <- cytof_dat %>% filter(timePoint %in% c('Week 8','Post Run-in'))



```


## P-values

Pairwise Wilcoxon test was used for each time point comparing responders vs non-responders. And paired Wilcoxon test was used for baseline vs post run-in and post run-in vs week 8.
We took the differences between all time points and then compare those differences between  responders vs non-responders to see if there is significant difference in changes after treatment

```{r  fig.width=6}

# Calculate p-values from Wilcoxon test: paired for time points and not paired for response
res = c()
# split by clusters
cytof_dat_byClust = split(cytof_dat, f = factor(cytof_dat$cluster))
for(j in names(cytof_dat_byClust))
{
  #
  d = cytof_dat_byClust[[j]]
  for (k in markers)
  {
    # create per patient matrix with three time points in columns
    # some values would be NA
    mat <- data.frame(matrix(nrow = length(unique(d$patientID)),
      ncol = 3, dimnames = list(unique(d$patientID),
                      levels(factor(d$timePoint)))), check.names = F)
    # fill in matrix
    for(i in levels(factor(d$timePoint)))
    {
      d1 = d %>% filter(timePoint == i)
      mat[d1$patientID, i] = d1[,k]
    }
    # add response
    mat$RECIST = cytof_dat[rownames(mat),"RECIST"]
    # create treatment effect 
    # substract baseline from post run-in and week 8 
    mat$delta_T1_B =  mat[,2]-mat[,1]
    mat$delta_T2_B =  mat[,3]-mat[,1]
    # and week 8 from post run-in 
    mat$delta_T2_T1 =  mat[,3]-mat[,2]
    
    # running paired test
    # Baseline vs post run-in
    p1 = wilcox.test(mat[,1], mat[,2], paired = T)$p.value
    # post run-in vs week 8
    p2 = wilcox.test(mat[,2], mat[,3], paired = T)$p.value
    
    # test by response
    # for baseline 
    p3 = wilcox.test(Baseline ~ RECIST, data = mat)$p.value
    # for post run-in 
    p4 = wilcox.test(mat[,2] ~ mat[,"RECIST"], data = mat)$p.value
    # for week 8
    p5 = wilcox.test(mat[,3] ~ mat[,"RECIST"], data = mat)$p.value
    # compare differences in treatment effect in responders vs non-responders
    p6 = wilcox.test(delta_T1_B ~ RECIST, data = mat)$p.value
    p7 = wilcox.test(delta_T2_B ~ RECIST, data = mat)$p.value
    p8 = wilcox.test(delta_T2_T1 ~ RECIST, data = mat)$p.value
   
    # combine all p-values and add means
    res = rbind(res, c(marker = k, cluster = j, BvsT1 = round(p1,3),
                       T1vsT2 = round(p2,3), B_RvsNR = round(p3,3), 
                       T1_RvsNR = round(p4,3), T2_RvsNR = round(p5,3), 
                       d_T1_B_RvsNR = round(p6,3), 
                       d_T2_B_RvsNR = round(p7,3), 
                       d_T2_T1_RvsNR = round(p8,3),
                       mean_d_T1_B_R = round(mean(mat %>% filter(RECIST == "Responder") %>% select(delta_T1_B) %>% unlist(), na.rm = T),3),
                       mean_d_T1_B_NR = round(mean(mat %>% filter(RECIST == "Non-responder") %>% select(delta_T1_B) %>% unlist(), na.rm = T),3),
                       mean_d_T2_B_R = round(mean(mat %>% filter(RECIST == "Responder") %>% select(delta_T2_B) %>% unlist(), na.rm = T),3),
                       mean_d_T2_B_NR = round(mean(mat %>% filter(RECIST == "Non-responder") %>% select(delta_T2_B) %>% unlist(), na.rm = T),3),
                       mean_d_T2_T1_R = round(mean(mat %>% filter(RECIST == "Responder") %>% select(delta_T2_T1) %>% unlist(), na.rm = T),3),
                       mean_d_T2_T1_NR = round(mean(mat %>% filter(RECIST == "Non-responder") %>% select(delta_T2_T1) %>% unlist(), na.rm = T),3)))
  }
}
res = res[order(res[,'marker']),]

datatable(res)

write.csv(res,file = "table_cytof_wilcox_all_markers.csv", row.names = F)


# make boxplots for markers/clusters that have p-value < 0.05 in d_T1_B_RvsNR
sets = data.frame(res)
sets = sets %>% filter(d_T1_B_RvsNR < 0.05 & cluster%in% c('B','CD4_T','CD8_T','DNT','DPT','NK_like',"cDC1"))

for (i in 1:nrow(sets))
{
  k = sets[i,'cluster']
  j = sets[i,'marker']
d = cytof_dat %>% filter(cluster == k)
  df = cbind(protein = d[,j], 
             d[,c("patientID", "timePoint", "RECIST","Subtype") ] )

    p = ggpaired(df , x="timePoint",y="protein", id = "patientID", 
                 line.size = 0.4, title = paste0(j,". ", k), 
                 line.color = "RECIST",palette =  recistCol, xlab="Time") +
      facet_wrap("RECIST") +
      scale_x_discrete(labels=timeLab)
    
    print(p)
}

# additionally, plot CD103 in cDC1 cluster

  k = "cDC1"
  j = "CD103"
d = cytof_dat %>% filter(cluster == k)
d = CreateAllFacet(d,"RECIST")
  df = cbind(protein = d[,j], 
             d[,c("patientID", "timePoint", "RECIST","Subtype", "facet") ] )

    p = ggpaired(df , x="timePoint",y="protein", id = "patientID", 
                 line.size = 0.4, title = paste0(j,". ", k), 
                 line.color = "RECIST",palette =  recistCol, xlab="Time") +
      facet_wrap("facet") +
      scale_x_discrete(labels=timeLab)
    
    print(p)

```


## Baseline vs Post Run-in 

```{r  fig.width = 10, fig.height=15}
#ggpaired(df_BvT1 %>% filter(cluster %in% c("GMDSC")),
for (i in markers)
{
  p = ggpaired(df_BvT1, x="timePoint",y=i, id = "patientID", line.color = "RECIST", line.size = 1,
   palette =  recistCol, title = paste0(i,". Baseline vs Post Run-in"), xlab="Time", ylab="MMI") + 
    scale_x_discrete(labels=timeLab)+ 
    facet_wrap("cluster", scales = 'free') +
    theme(axis.text.x=element_text(angle = 45, hjust = 1))

  print(p)
}


```

## Post Run-in vs Week 8

```{r  fig.width = 10, fig.height=15}

for (i in markers)
{
  p = ggpaired(df_T1vT2, x="timePoint",y=i, id = "patientID", line.color = "RECIST", line.size = 1,
   palette =  recistCol, title = paste0(i,". Post Run-in vs Week 8"), xlab="Time", ylab="MMI") +
#    scale_x_discrete(labels=timeLab)+
    facet_wrap("cluster", scales = 'free') +
    theme(axis.text.x=element_text(angle = 45, hjust = 1))

  print(p)
}

```

```{r}
sessionInfo()
```