---
title: "J15221 Evanthia TNBC - CyTOF marker expression"
author: "Ludmila Danilova & Joseph Tandurella"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE)
rm(list = ls())

library(dplyr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(data.table)
library("DT")
library(gridExtra)
library(scales)

source("functions.r")
```

Create figures for the clinical paper

All data is coming from tables for the paper


```{r fig.width = 10}
# labels for time points
timeLab  = c("Baseline", "Post Run-in", "Week 8")

# set colors
recistCol = c("Non-responder" = "red","Responder" = "blue")
cbrCol = c("CR,PR,SD" = "blue","PD" = "red")
subtypeCol = c("HR+" = "red", "TNBC" = "blue")
timeCol = setNames(c("black", 'grey45', 'grey75'), timeLab)

# set root folder
rootDir = "./"

# Load data patient data
dta_pts = read.csv(paste0(rootDir,"input_tables/table_24patient_info.csv"), row.names = 1)
dta_pts$ID = rownames(dta_pts)

# remove patients with no response
dta_pts = dta_pts %>% filter(Best_Resp != "Unknown")
dim(dta_pts)
table(dta_pts$RECIST)

# load cluster proportions
dat = read.csv(paste0(rootDir,"input_tables/table_per_sample_values.csv"))
props = grep("proportion", colnames(dat), value = T)

#===========================
# add patient info 
dat = cbind(dat,  dta_pts[dat$patientID,])
```

# Cluster proportions
## Table with p-values
Pairwise Wilcoxon test was used for each time point comparing responders vs non-responders. And paired Wilcoxon test was used for baseline vs post run-in and post run-in vs week 8. We took the differences between all time points and then compare those differences between responders vs non-responders to see if there is significant difference in changes after treatment

```{r fig.width = 10}

prop_res = c()
for (k in props)
{
  prop_res = rbind(prop_res, c(cluster = gsub("_proportion","",k), runWilcox(k, dat, dta_pts)))
}

knitr::kable(prop_res, "simple",digits = 3)

write.csv(prop_res,file = "table_cytof_wilcox_props.csv", row.names = F)


```

# Marker expression by clusters


```{r fig.width = 10}

#==========================
# file with marker expression
cytof_dat = read.csv(paste0(rootDir,"input_tables/table_cytof_marker_expression.csv"))


markers = colnames(cytof_dat)[4:ncol(cytof_dat)]
#===========================
# add patient info 
cytof_dat = cbind(cytof_dat,  dta_pts[cytof_dat$patientID,])
# save data to submit with the paper
write.csv(cytof_dat, file = "Source_data_figure4_EDF2_EDF3.csv", row.names = F)


cat("The number of patients with CYTOF data:", length(intersect(cytof_dat$patientID, dta_pts$ID)))
# print the number of pairs between time points
mat = printPairs(cytof_dat)

# take only baseline and time 1 values
df_BvT1 <- cytof_dat %>% filter(timePoint %in% c('Baseline','Post Run-in'))

# take only time 1 and time 2 values
df_T1vT2 <- cytof_dat %>% filter(timePoint %in% c('Week 8','Post Run-in'))



```


## Table with p-values

Pairwise Wilcoxon test was used for each time point comparing responders vs non-responders. And paired Wilcoxon test was used for baseline vs post run-in and post run-in vs week 8.
We took the differences between all time points and then compare those differences between  responders vs non-responders to see if there is significant difference in changes after treatment

```{r  fig.width=6}

# Calculate p-values from Wilcoxon test: paired for time points and not paired for response

res = c()
# split by clusters
cytof_dat_byClust = split(cytof_dat, f = factor(cytof_dat$cluster))

for(j in names(cytof_dat_byClust))
{
  #
  d = cytof_dat_byClust[[j]]
  for (k in markers)
  {
    res = rbind(res, c(marker = k, cluster = j, runWilcox(k, d, dta_pts)))
  }
}

res = res[order(res[,'marker']),]

datatable(res)

write.csv(res,file = "table_cytof_wilcox_all_markers.csv", row.names = F)

```
## Boxplot (Responders and Non-responders)

```{r  fig.width=6}

# make boxplots for markers/clusters that have p-value < 0.05 in d_T1_B_RvsNR
sets = data.frame(res)
sets = sets %>% filter(d_T1_B_RvsNR < 0.05 & cluster%in% c('B','CD4_T','CD8_T','DNT','DPT','NK_like',"cDC1"))

for (i in 1:nrow(sets))
{
  k = sets[i,'cluster']
  j = sets[i,'marker']
d = cytof_dat %>% filter(cluster == k)
  df = cbind(protein = d[,j], 
             d[,c("patientID", "timePoint", "RECIST","Subtype") ] )

    p = ggpaired(df , x="timePoint",y="protein", id = "patientID", 
                 line.size = 0.4, title = paste0(j,". ", k), 
                 line.color = "RECIST",palette =  recistCol, xlab="Time") +
      facet_wrap("RECIST") +
      scale_x_discrete(labels=timeLab)
    
    print(p)
}

# additionally, plot CD103 in cDC1 cluster

  k = "cDC1"
  j = "CD103"
d = cytof_dat %>% filter(cluster == k)
d = CreateAllFacet(d,"RECIST")
  df = cbind(protein = d[,j], 
             d[,c("patientID", "timePoint", "RECIST","Subtype", "facet") ] )

    p = ggpaired(df , x="timePoint",y="protein", id = "patientID", 
                 line.size = 0.4, title = paste0(j,". ", k), 
                 line.color = "RECIST",palette =  recistCol, xlab="Time") +
      facet_wrap("facet") +
      scale_x_discrete(labels=timeLab)
    
    print(p)

```


## Boxplot (Baseline vs Post Run-in)

```{r  fig.width = 10, fig.height=15}
#ggpaired(df_BvT1 %>% filter(cluster %in% c("GMDSC")),
for (i in markers)
{
  p = ggpaired(df_BvT1, x="timePoint",y=i, id = "patientID", line.color = "RECIST", line.size = 1,
   palette =  recistCol, title = paste0(i,". Baseline vs Post Run-in"), xlab="Time", ylab="MMI") + 
    scale_x_discrete(labels=timeLab)+ 
    facet_wrap("cluster", scales = 'free') +
    theme(axis.text.x=element_text(angle = 45, hjust = 1))

  print(p)
}


```

## Boxplot (Post Run-in vs Week 8)

```{r  fig.width = 10, fig.height=15}

for (i in markers)
{
  p = ggpaired(df_T1vT2, x="timePoint",y=i, id = "patientID", line.color = "RECIST", line.size = 1,
   palette =  recistCol, title = paste0(i,". Post Run-in vs Week 8"), xlab="Time", ylab="MMI") +
#    scale_x_discrete(labels=timeLab)+
    facet_wrap("cluster", scales = 'free') +
    theme(axis.text.x=element_text(angle = 45, hjust = 1))

  print(p)
}

```

```{r}
sessionInfo()
```